Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-04-14 paper/paper.Rmd
supplement update
prep for submission

Yet given the logistical hurdles involved in sampling from this population, individual experiments measuring processing speed with young children typically recruit relatively small samples in a restricted range of ages. These samples provide neither the breadth of ages nor the number of participants needed to estimate how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children). To overcome this limitation, we created Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank]. This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements during real-time language processing. This resource allows us to gain an unprecedented view of the development of word recognition across a large sample of children.

@@

We began by examining developmental changes in children's word recognition. Figure \ref{fig:timecourse} depicts the average timecourse at different ages across all datasets. Intuitively, these timecourses show gradual increases in accuracy (higher overall proportion target looking) and speed (faster looking to the target after hearing a label) as age increases. To systematically characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms). For each window, we averaged all fixations within the window to compute a continuous proportion of target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial. In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison). Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analyses.

@@

To investigate within-person relationships between LWL and vocabulary, we began by fitting longitudinal growth models to the portion of the data containing multiple LWL sessions for individual children. We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study. We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset). This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`). This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time, confirming that findings of @fernald2012 were robust across the full set of longitudinal data in our dataset.

@@

On the other hand, it is possible that differences in predicted growth trajectories are due to coupling between vocabulary size and language processing across the entire developmental period, rather than a predictive relationship specifically between $t_0$ RT and vocabulary growth. To test this relationship, we used longitudinal structural equation models. We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months from the initial measurement (i.e., $t_0, ..., t_4$) and fit individual growth across each of these variables. We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset. (We present a model of coupled growth in the observed variables here for simplicity; see SI: Latent Factor Growth for a comparable model using growth in the latent factors). The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}. Overall fit statistics were generally acceptable (Confirmatory fit index: `r round(slopes_long_summary$fit["cfi"], 3)`, RMSE: `r round(slopes_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(slopes_long_summary$fit["rmsea.pvalue"], 3)`).

@@

We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms). We also removed RTs shorter than 367 ms, as these were unlikely to be generated based on the specific linguistic stimulus. We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements. At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 2)`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 2)`% missingness for long window accuracies.

>>>>

Yet given the logistical hurdles involved in sampling from this population, individual experiments measuring processing speed with young children typically recruit relatively small samples in a restricted range of ages. These samples provide neither the breadth of ages nor the number of participants needed to estimate how word recognition changes developmentally and how it connects with other aspects of early language development (see [@colby2023efficiency; @mcmurray2022slow] for examples of these analyses in school-aged children). To overcome these limitations, we created Peekbank, an open database of LWL data from young children, stored in a harmonized format [@zettersten2023peekbank]. This dataset unifies and carefully curates a large amount of eye-tracking data from studies with infants and toddlers, representing cumulatively over 12 million individual samples of children's eye movements during real-time language processing. The Peekbank dataset allows us to gain an unprecedented view of the development of word recognition across a large sample of children.

@@

We began by examining developmental changes in children's word recognition. Figure \ref{fig:timecourse} depicts the average timecourse at different ages across all datasets (not controlling for any variation in items and procedures across age groups). Intuitively, these timecourses show gradual increases in accuracy (higher overall proportion target looking) and speed (faster looking to the target after hearing a label) as age increases. To characterize age gradients in speed and accuracy across children, we computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms). For each window, we averaged all fixations within the window to compute a continuous proportion of target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial. In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study (modeling item and procedural) variation across studies) and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison). Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analyses.

@@

To investigate within-person relationships between LWL and vocabulary, we began by fitting longitudinal growth models to the portion of the data containing multiple LWL sessions for individual children. We first reproduced the analysis reported in @fernald2012, in which between-person differences in longitudinal growth in productive vocabulary were predicted based on between-person differences in speed during the initial session of the study. We fit a mixed effects model predicting growth in vocabulary as a quadratic function of age, RT at study initiation ($t_0$), and their interaction (as well as random effects of age by participant and by dataset). This model revealed a significant effect of $t_0$ RT (`r papaja::apa_print(growth_mod)$full_result$rt_t0`) and an interaction between $t_0$ RT and the quadratic age predictor (`r papaja::apa_print(growth_mod)$full_result$polyage_15_22_rt_t0`). This analysis suggests that children with faster initial RTs show both larger vocabularies and faster vocabulary growth over time.

@@

On the other hand, it is possible that differences in predicted growth trajectories are due to coupling between vocabulary size and language processing across the entire developmental period, rather than a predictive relationship specifically between $t_0$ RT and vocabulary growth. To test this relationship, we used longitudinal structural equation models. We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months from the initial measurement (i.e., $t_0, ..., t_4$) and fit individual growth across each of these variables. We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset. The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}. Overall fit statistics were generally acceptable (Confirmatory fit index: `r round(slopes_long_summary$fit["cfi"], 3)`, RMSE: `r round(slopes_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(slopes_long_summary$fit["rmsea.pvalue"], 3)`).

@@

We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms). We also removed RTs shorter than 367 ms, as these were unlikely to be generated based on the specific linguistic stimulus. We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements. At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 3)*100`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 3)*100`% missingness for long window accuracies.

### 2025-04-14 paper/supplemental_information.Rmd
prep for submission

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                ~ age_scale(.x, age))) 
fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
            use = "pairwise.complete.obs")
```

```{r}
fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
         use = "pairwise", rotate = "varimax")

as.data.frame(unclass(loadings(fs))) |> 
  kable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
```

@@

xtable(anova(fa1_fit, fa2_fit, fa2b_fit, fa2v_fit, fa3_fit), digits = 2)
```
<!-- xtable 1.8-4 package -->
<!-- % Fri Apr  4 15:15:52 2025 -->
\begin{table}[ht]
\centering
\caption{Model comparison for alternative factor structures. $p$-values show differences between adjacent models; no $p$-values are shown for comparisons between non-nested models. \label{tab:modcomp}}
\begin{tabular}{lrrrrrrrr}
  \hline
 & Df & AIC & BIC & Chisq & Chisq diff & RMSEA & Df diff & Pr($>$Chisq) \\ 
  \hline
Three-factor  & 6 & 46346.69 & 46475.43 & 91.96 &  &  &  &  \\ 
  Two-factor (vocab)  & 8 & 46397.74 & 46514.22 & 147.01 & 55.05 & 0.09 & 2 & $<$ 0.0001 \\ 
  Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
  Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0.00 &  \\ 
  One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

Figure 3 shows the result of a parallel analysis supporting the presence of three factors in the exploratory factor analysis. Table \ref{tab:loadings} shows the factor loadings for the exploratory three-factor solution using varimax rotation. The first factor is primarily driven by vocabulary measures, the second by reaction time, and the third by accuracy measures.

```{r parallel, fig.cap="Parallel analysis scree plot showing the eigenvalues for each factor, for actual, simulated, and resampled data."}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                ~ age_scale(.x, age))) 
outputs <- capture.output(fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
            use = "pairwise.complete.obs"))
```

```{r eval=FALSE}
fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
         use = "pairwise", rotate = "varimax")

as.data.frame(unclass(loadings(fs))) |> 
  xtable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
```

<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
<!-- % Mon Apr 14 12:25:14 2025 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
 & F1 & F2 & F3 \\ 
  \hline
RT & -0.19 & 0.81 & -0.30 \\ 
  RT var & -0.10 & 0.81 & -0.22 \\ 
  long window accuracy & 0.33 & -0.31 & 0.55 \\ 
  long window accuracy var & -0.10 & 0.26 & -0.65 \\ 
  production vocabulary & 0.95 & -0.04 & 0.30 \\ 
  comprehension vocabulary & 0.61 & -0.16 & -0.01 \\ 
  age & 0.63 & -0.14 & 0.37 \\ 
   \hline
\end{tabular}
\caption{\label{tab:loadings}Factor loadings for the exploratory three factor solution using varimax rotation.} 
\end{table}

@@

xtable(anova(fa1_fit, fa2_fit, fa2b_fit, fa2v_fit, fa3_fit), digits = 2)
```
<!-- xtable 1.8-4 package -->
<!-- % Fri Apr  4 15:15:52 2025 -->
\begin{table}[ht]
\centering
\caption{Model comparison for alternative factor structures. $p$-values show differences between adjacent models; no $p$-values are shown for comparisons between non-nested models. \label{tab:modcomp}}
\begin{tabular}{lrrrrrrrr}
  \hline
 & Df & AIC & BIC & Chisq & Chisq diff & RMSEA & Df diff & Pr($>$Chisq) \\ 
  \hline
Three-factor  & 6 & 46346.69 & 46475.43 & 91.96 &  &  &  &  \\ 
  Two-factor (vocab)  & 8 & 46397.74 & 46514.22 & 147.01 & 55.05 & 0.09 & 2 & $<$ 0.0001 \\ 
  Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
  Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0 &  \\ 
  One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 
   \hline
\end{tabular}
\end{table}
